---
title: "Municipalities Should Prioritize Preparation for Tornadoes and Floods"
output: 
  html_document:
    keep_md: true
---

**Author - Harland Hendricks**

National Weather Service data was used to determine which types of weather events cause the most harm to the population and the most economic damage to property and crops.  This analysis will inform government and municipal managers in the prioritization of resources to prepare for weather events.  The analysis showed that tornadoes, by a wide margin, cause the most harm to the population when taking into account total injured and dead due to an event.  The analysis also showed that the most economic damage is caused by flooding, hurricanes/typhoons, and tornadoes.  I recommend that local, state, and federal governments prioritize resources for tornado education, warning, and protection to best protect the population.  I also recommend that these governments prioritize resources to improve structural integrity of buildings when faced with high winds from tornadoes and hurricanes/typhoons.  Finally, governments should work with engineers and focus on storm water management to protect locales from flooding damage to crops and property.    

## Data Processing

```{r}
sessionInfo()
```

**The following links are helpful to understand the data**

1. National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

2. National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

Set your desired working directory

Load required libraries

```{r}
library(R.utils)
library(dplyr)
library(ggplot2)
```

Load Data

```{r cache = TRUE}
fileurl = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if (!file.exists("data")) {
        dir.create("data")
        }
download.file(fileurl, destfile = "./data/stormdata.csv.bz2", 
                mode = "wb")
bunzip2("./data/stormdata.csv.bz2", "./data/stormdata.csv", remove = FALSE, skip = TRUE)
data <- read.csv("./data/stormdata.csv")
```

### Tidy data

Create a dataset for the analysis of types of weather events and affect on population health.

```{r}
healthdata <- data %>%
        select(EVTYPE, FATALITIES, INJURIES) %>%
        mutate(total_affected = FATALITIES + INJURIES) %>%
        rename_all(tolower)
head(healthdata, 20)
```

Create a dataset for the analysis of types of weather events and economic damage.

```{r}
econdata <- data %>%
        select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
        rename_all(tolower) %>%
        mutate(propdmgmult = propdmgexp) %>%
        mutate(cropdmgmult = cropdmgexp)
## Change columns to character
econdata$propdmgexp <- as.character(econdata$propdmgexp)
econdata$cropdmgexp <- as.character(econdata$cropdmgexp)
econdata$propdmgmult <- as.character(econdata$propdmgmult)
econdata$cropdmgmult <- as.character(econdata$cropdmgmult)
## Replace H, K, M, B with multiple
econdata <- econdata %>%
        mutate(propdmgmult = replace(propdmgmult, propdmgexp == "H", "100")) %>%
        mutate(propdmgmult = replace(propdmgmult, propdmgexp == "K", "1000")) %>%
        mutate(propdmgmult = replace(propdmgmult, propdmgexp == "M", "1000000")) %>%
        mutate(propdmgmult = replace(propdmgmult, propdmgexp == "B", "1000000000"))
## Characterize multiple column as numeric
econdata$propdmgmult <- as.numeric(econdata$propdmgmult)
econdata$cropdmgmult <- as.numeric(econdata$cropdmgmult)
## Add total columns to account for the multiple
econdata <- econdata %>%
        mutate(property_damage = propdmg * propdmgmult) %>%
        mutate(crop_damage = cropdmg * cropdmgmult) %>%
        select(evtype, property_damage, crop_damage)
econdata$total_damage <- rowSums(econdata[, c(2, 3)], na.rm = TRUE)
head(econdata, 20)
```

## Questions for Analysis

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Group healthdata by evtype and summarize by using the sum of the total_affected column.  Arrange total in descending order and filter by the top 5 events that are most harmful.

```{r}
health_summary <- healthdata %>%
        group_by(evtype) %>%
        summarize(total = sum(total_affected)) %>%
        as.data.frame(health_summary) %>%
        filter(total != 0) %>%
        arrange(desc(total)) %>%
        slice(1:5)
health_summary
```

2. Across the United States, which types of events have the greatest economic consequences?

Group econdata by evtype and summarize by using the sum of the total_damage column.  Arrange total in descending order and filter by the top 5 events that are most economically damaging.

```{r}
econ_summary <- econdata %>%
        group_by(evtype) %>%
        summarize(total = sum(total_damage)) %>%
        as.data.frame(econ_summary) %>%
        filter(total != 0) %>%
        arrange(desc(total)) %>%
        slice(1:5)
econ_summary
```

## Results

### Figure 1

```{r}
ggplot(health_summary, aes(reorder(evtype, -total), total)) + geom_col() + labs(title =
        "Total Injuries and Deaths by Top 5 Weather Events") + labs(x = "Weather Event Type") +
        labs(y = "Total Injuries and Deaths")
```

The graph in figure 1 shows that the top three weather events harmful to the population are:

        1.  Tornado (By a wide margin)
        2.  Excessive Heat
        3.  Thunderstorm Wind

### Figure 2

```{r}
options(scipen = 999)
ggplot(econ_summary, aes(reorder(evtype, -total), total)) + geom_col() + labs(title =
        "Total Damage by Top 5 Weather Events") + labs(x = "Weather Event Type") +
        labs(y = "Total Crop and Property Damage (USD)")
options(scipen = 99)
```

The graph in figure 2 shows that the top three weather events that cause the most economic damage are:

        1.  Flood
        2.  Hurricane/Typhoon
        3.  Tornado

It should also be noted that 3/5 of these weather event types are flood/water related.

## Recommendations

This analysis shows that municipalities should prioritize resources as follows:

1.  To protect the population from weather events, the municipality should educate the public on tornado and thunderstorm actions to reduce the likelihood of injury or death.  The municipality should also provide tornado/thunderstorm warning systems and shelters.  Lastly, the municipality should provide a clean water supply to hydrate and reliable power for air conditioning to reduce heat related injuries and deaths.

2.  To protect the municipality economically, the municipality should work with engineers to improve storm water management to reduce damage from flooding.  Also, the municipality should ensure that engineering specifications are sufficient to protect property from high winds associated with hurricanes, typhoons, and tornadoes. 
